Aneuploidy may be an evolutionary sidetrack on the path to adaptation

### 2022-07-26 ms/ms.tex
revised FA with deterministic dynamics, log-log in supp

\paragraph{Estimated rates and fitness effects of aneuploidy and mutation.} 
We inferred the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
\Cref{fig:ppc-plot}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.

We tested the \emph{transient adaptive solution hypothesis} by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:ppc-plot}A). 
This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:ppc-plot}A).

This result is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:ppc-plot}B). 
The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.  
Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:ppc-plot}C and \Cref{fig:posterior_FA_supp}).

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Simulated genotype frequency dynamics.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Shown are 50 simulations using the MAP estimate parameters.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
     }
  \label{fig:ppc-plot}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later. 
Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:ppc-plot}).

@@

% FIG posterior F_A supplementary
\begin{figure}[p]
	\begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_1.pdf}      
	\end{subfigure}
	  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_2.pdf}
  \end{subfigure}
	\caption{\textbf{Posterior distribution of $F_A$.}
	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:ppc-plot} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
	}
    \label{fig:posterior_FA_supp}
\end{figure}

>>>>

\paragraph{Estimated rates and fitness effects of aneuploidy and mutation.} 
We inferred the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets. See \emph{Supporting material} for sensitivity analysis.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.965\cdot10^{-6}\ [2.718\cdot10^{-7}-3.589\cdot10^{-6}]$, corresponds to a mutation target size of $\sim 10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$ (ref.~\citep{Zhu2014}) or $3.3\cdot10^{-10}$ (ref.~\citep{Lynch2008}).
The estimated aneuploidy rate, $\delta=1.72\cdot10^{-3}\ [1.47\cdot10^{-3}-2.786\cdot10^{-3}]$ is much higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimated $6.7\cdot10^{-6}$ chromosome gain events per generation, and \citet{Kumaran2013} estimate $3.0-4.3\cdot10^{-5}$ chromosome loss events per generation (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the cost of trisomy is $c = w_{\eumt}-w_{\anmt} = 0.003$ (or 0.3\%) and the benefit of trisomy is $w_{\anwt}-1-c=0.019$ (1.9\%), whereas the benefit of beneficial mutation is $w_{\eumt}-1=0.028$ (2.8\%).

@@

We simulated 50 replicate genotype frequency dynamics using the MAP estimate parameters. 
\Cref{fig:FA}A shows the simulated frequencies of the four genotypes (\euwt, \anwt, \anmt\ and \eumt), as well as the frequencies of \eumt\ cells that arose from either \anwt\ cells via a sequences of mutation and chromosome loss events ($2n^*_A$), or directly from \euwt\ cells via a mutation event (\eumtM).
We find that \anmt\ never reaches substantial frequency as it is quickly replaced by \eumt\ in a process similar to \emph{stochastic tunneling}~\citep{Iwasa2004,Komarova2003}.

We tested the \emph{transient adaptive solution hypothesis} by estimating $F_A$, the expected frequency of \eumt\ that arose from \anwt, computed as the average frequency of such cells at the end of simulations using the MAP estimate parameters.
Surprisingly, we observe that the majority of \eumt\ cells are a product of a direct mutation in \euwt\ cells, rather than descending from \anwt\ cells ($F_A^{MAP} = 0.106$, \Cref{fig:FA}A). 
This is despite the fact that the \anwt\ genotype reaches high frequencies in the population (at least 0.98, \Cref{fig:FA}A).

This result is not unique to the MAP parameter estimate. 
We simulated genotype frequency dynamics using parameter samples from the posterior distribution (\Cref{fig:posterior}), and computed the posterior distribution of $F_A$ (\Cref{fig:FA}B). 
The mean $F_A$ was just 0.1673 [0.0154-0.370 95\% CI] and only in 489 of 100,000 posterior samples (0.489\%) $F_A$ was  larger than 0.5, that is, \eumt\ was more likely to arise from an aneuploid cell.  
Thus, if we sample a random cell from the evolved \eumt\ population, it is more likely to have descended directly from an euploid cell than from an aneuploid cell.
The probability of \eumt\ descending from \anwt\ ($F_A$) increases with the aneuploidy rate, $\delta$, and decreases with the mutation rate, $\mu$, and in some cases can also be affected by the fitness parameters (\Cref{fig:FA}C and \Cref{fig:posterior_FA_supp}).

@@

% Fig dynamics
\begin{figure}[h!]
      \centering
      \includegraphics[width=\textwidth]{../figures/FA.pdf}      
  \caption{
    \textbf{Predicted frequency of aneuploid-descended cells.}     
    \textbf{(A)} Posterior predicted genotype frequencies over time, including the source of \eumt: $2n^*_A$ arose from $2n+1$, whereas $2n^*_M$ arose directly from $2n$. Colored curves are 50 simulations using the MAP estimate parameters. Black dashed curves are the expected genotype frequencies without genetic drift. 
     See \Cref{fig:MAP_freq_dynamics_log} for log-log scale, in which the sequence of events is easier to observe.
    \textbf{(B)} Posterior distribution of $F_A$, the expected frequency of \eumt\ cells descended from \anwt\ cells, computed as the average frequency at the end of 100 simulations for 100,000 samples from the parameter posterior distribution.
    Solid and dashed lines show the mean and 95\% CI.
	\textbf{(C)} $F_A$ values (color coded) from panel B, with their corresponding mutation rate $\mu$ on x-axis and aneuploidy rate $\delta$ on the y-axis.
	White star shows the MAP estimate. See also \Cref{fig:posterior_FA_supp}.
	\textbf{(D)} $F_A$ as a function of the population size, $N$, in posterior predictions with MAP parameters. Markers show $F_A$ in 250 simulations per population size. Error bars show mean $F_A$ with 95\% CI (bootstrap, $n=10,000$). Vertical dashed line for population size in the experiment, $6.425\cdot10^6$. Horizontal line for $F_A^{MAP}=0.106$. 
     }
  \label{fig:FA}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

If indeed the evolved euploid population is descended from the aneuploid population, then mutant alleles that were common in the aneuploid populations should also be common in the evolved euploid population.
However, we found that this is not the case (\Cref{fig:mutation_frequencies}): mutant allele frequencies in the aneuploid and euploid populations are negatively correlated, such that common alleles in the former are rare in the later. 
Furthermore, we developed an evolutionary genetic model of adaptive evolution by aneuploidy and mutation (\Cref{fig:models}), fitted it to the experimental results of \citet{Yona2012}, and used it to predict the genotype frequency dynamics.
The model predicted that only about 10\% of the evolved euploid population descended from aneuploid cells--that is, the majority of the euploid population are not descended from aneuploid cells, but rather are direct descendants of the ancestral population (\Cref{fig:FA}).

@@

% FIG MAP frequency dynamics in log scale
\begin{figure}[p]
    \centering
	\includegraphics[width=\textwidth]{../figures/MAP_freq_dynamics.pdf}
	\caption{\textbf{Posterior predicted genotype frequencies in log-log scale.}
	Frequency dynamics of the different genotypes with MAP parameter estimates, same as \Cref{fig:FA}A, but in log-log scale. Black dashed curves for a deterministic model without genetic drift. 
	Clearly, appearance of \anwt\ and \eumtM\ is deterministic.
	Appearance of \anmt, and therefore \eumtA, is stochastic, but their frequency dynamics after reaching frequency of roughly 0.001 is deterministic. 
	}
    \label{fig:MAP_freq_dynamics_log}    
\end{figure}

% FIG posterior F_A supplementary
\begin{figure}[p]
	\begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_1.pdf}      
	\end{subfigure}
	  \begin{subfigure}{0.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/F_A_delta-mu-ratio_2.pdf}
  \end{subfigure}
	\caption{\textbf{Posterior distribution of $F_A$.}
	\textbf{(A,B)} $F_A$ values (color coded) as in \Cref{fig:FA} for different parameter choices on the x- and y-axes. White star denotes the MAP estimate.
	}
    \label{fig:posterior_FA_supp}
\end{figure}
